Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-08 article.Rmd
Describe SSA dataset

The Social Security Administration provides annual data on the names and genders of babies stretching back to the year 1880. This is database encompases 91,320 unique names. For comparison, [example of other databases and their size]. What makes it even more useful, however, is its temporal dimension. This allows a researcher to figure out not just the likelihood that “Leslie” is a male or female name, but the likelihood that “Leslie” is a male or female name for a specific range of years, or for a specific year.

Using for example the SSA data for a few names (once it has been munged into a more useful form). The columns male and female record the number of people of that gender (sex?) born in that year who received Social Security cards.

name         year    female     male
---------  ------  --------  -------
jane         1890       372        0
jane         1990       771        8
john         1890        56     8502
john         1990        81    29066
madison      1890         0       14
madison      1990      1407      145

Furthermore, we can calculate the proportions of male and female and on that basis guess the gender for a given year.

name         year    female     male    proportion_male    proportion_female  gender  
---------  ------  --------  -------  -----------------  -------------------  --------
jane         1890       372        0             0.0000               1.0000  female  
jane         1990       771        8             0.0103               0.9897  female  
john         1890        56     8502             0.9935               0.0065  male    
john         1990        81    29066             0.9972               0.0028  male    
madison      1890         0       14             1.0000               0.0000  male    
madison      1990      1407      145             0.0934               0.9066  female
ADD LESLIE AND ANOTHER EXAMPLES WE USE

Furthermore, one can specify a range of years and calculate the proportion for that time period.

Given this data set, it is simple to perform perform a merge (a join, in SQL terms) on either a name or a combination of name and year to link this data to a date set.

>>>>

The Social Security Administration "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[Social Security Administration, "[Baby Names from Social Security Card Applications-National Level Data](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)," Data.gov dataset ID US-GOV-SSA-338 (accessed August 31, 2014). The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910. Social Security Administration, "[Baby Names from Social Security Card Applications-Data by State and District of Columbia](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-data-by-state-and-district-of-)," Data.gov dataset ID US-GOV-SSA-339 (accessed August 31, 2014).] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA has included any name which was used more than five times in a given year; otherwise the dataset is a 100 percent sample. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[CITE history of SSA. On the history of birth registration see Shane Landrum, "The State's Big Family Bible: Birth Certificates, Personal Identity, and Citizenship in the United States, 1840--1950" (PhD dissertation, Brandeis University, 2014). It is well-known among American historians that Social Security was written to exclude farmworkers with the aim of excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

TODO IPUMS
